---
title: "Covid Package"
author: "Dan Golden"
date: "12/13/2020"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
library(covidDataSummarizer)
```


## Overview

This report will cover the three focuses of this described package: pulling COVID data from the CDC, presenting example summary plots, and performing time series analysis.


## Centers for Disease Control and Prevention Data

The Centers for Disease Control (CDC) makes COVID-19 case and death data available to the public. This data can be accessed via a URL that allows subsets of the large dataset to be downloaded. This service is provided by an application programming interface (API). The API allows public access to COVID-19 data in either CSV or JSON formats. 

A CSV is a standard format for storing data that uses comma separated values. Data fields are separated by commas thus identifying unique entries. JSON is a newer data format that is more flexible than csv. Data for this project will be retrieved in CSV format. The CSV will be read into an R dataframe for use in analysis.

The CDC data can be retrieved using three different methods. The first method is simply to download an entire CSV that contains all available data. The second method uses simple additions to the base URL that allow for a minimal amount of data subsetting. The third method, which is the method focused on in this report, is the use of query language within the URL to subset data.

The API allowing access to CDC COVID-19 data is build using Socrata. The Socrata site includes the following definition of its service, "The Socrata Open Data API allows you to programmatically access a wealth of open data resources from governments, non-profits, and NGOs around the world."

Rather than downloading an entire dataset, or applying minimal subsetting, Socrata allows richer subsetting of COVID-19 data directly from the download URL. For example specific dataset fields can be selected, beginning and ending time periods can be defined, and data can be filtered to selected states. 



Field             | Datatype    | Description
-----             | --------    | -----------
submission_date   | timestamp   | Date and time record submited
state             | text        | State where record applies 
tot_cases         | number      | Total number of cases
conf_cases        | number      | Total confirmed cases
prob_cases        | number      | Total probable cases
new_case          | number      | Number of new cases
pnew_case         | number      | Number of new probable cases
tot_death         | number      | Total number of deaths
conf_death        | number      | Total number of confirmed deaths
prob_death        | number      | Total number of probable deaths
new_death         | number      | Number of new deaths
pnew_death        | number      | Number of new probable deaths


- United States COVID-19 Cases and Deaths by State over Time: https://dev.socrata.com/foundry/data.cdc.gov/9mfq-cb36


## Supplemental Datasets

Three additional datasets are included in the package. All three datasets were obtained from publicly accessible government sites. The datasets are included to support additional analysis and data aggregate to better understand the impact of COVID are various populations.

SAHIE Insurance Data
https://www.census.gov/data/datasets/time-series/demo/sahie/estimates-acs.html

Demographic Data (Race and Ethnicity)
https://covid19.census.gov/datasets/race-and-ethnicity-states/data

Demographic Data (Age and Sex)
https://covid19.census.gov/datasets/population-by-age-and-sex-states



```{r}
pull_covid_data(start_date="2020-06-01", end_date="2020-07-01")
```

```{r}
pull_covid_data(start_date="2020-06-01", end_date="2020-07-01")
```


## Plots

Example plots

```{r}

```


## Time Series

Example time series

```{r}

```



## Conclusion

Text
