---
title: "Covid Data Summarizer"
author: "Gerard Garvan, Dan Golden, Vineeth Kurnala"
date: "`r Sys.Date()`"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Covid Data Summarizer}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)

options(warn=-1)
library(dplyr)
library(lubridate)
library(zoo)
library(tidyr)
library(purrr)
library(timetk)
library(sweep)
library(ggplot2)
library(tidyquant)
library(forecast)
library(covidDataSummarizer)
```


## Retrieve Data

```{r}
# Example using dates
pull_covid_data(start_date="2020-06-01", end_date="2020-07-01", limit=5)
```

```{r}
# Example using dates and states
pull_covid_data(start_date="2020-06-01", end_date="2020-07-01", states=c("FL", "GA", "NY"), limit=5)
```

```{r}
# Example using dates and states and fields
pull_covid_data(start_date="2020-06-01", end_date="2020-07-01", 
                states=c("FL", "GA", "NY"), 
                fields=c("submission_date", "state", "tot_cases", "tot_death"), limit=10)
```

## Plots

The two plot functions covid_cases_graph(start_date,end_date) and covid_deaths_graph(start_date,end_date) give graphical summaries over the time specified.

covid_cases_graph(start_date,end_date) plots the tsix states that had the largest number of new cases at the end date( essentially x= date and y= number of new 
COVID cases).

Similarly covid_deaths_graph(start_date,end_date) plots the six states that had the largest number of new deaths at the end date (x= date and y=number of new deaths due to COVID). 

Both graph functions also have an option to report cases or deaths adjusting for population, i.e. cases per 100,000,000 people or deaths per 100,000,000 people. 

```{r}
#Example of covid_cases_graph
Thanksgiving2020<-"2020-11-19"
today<-Sys.Date()
covid_cases_graph(start_date = Thanksgiving2020,end_date = today,per100k = T) 
#example of covid_deaths_graph
covid_deaths_graph("2020-11-01","2020-12-01",per100k = T)
```

## Time Series Forecasting

**Defaults

* ```start_date = 2020-01-22``` (COVID-19 data in the CDC database is available from this date)
* ```state.filter = c("all")``` (state.filter uses 2 letter state codes or "all". If not specified, the function forecasts new cases or new deaths across entire USA)
* ```region.filter``` (If not specified, the function forecasts new cases or new deaths across entire USA i.e., the function uses ```state.filter=c("all")```)
* ```model``` (Uses ```auto.arima``` (ARIMA) model to forecast new cases or new deaths)
* ```pred.days = 30``` (If not specified, the funciton forecasts new cases or new deaths from now up to the next 30 days. The maximum limit for prediction is 100 days)

### Examples

* Forecasting new COVID-19 deaths across USA from now until next 30 days using data from Apr 30, 2020 by ARIMA model. 
```{r}
covid_ts_forecast(start_date='2020-04-30', fvar="new_death", model=c("auto.arima"), pred.days = 30)
```

* Forecasting new COVID-19 cases across USA from now until next 30 days using data from Apr 30, 2020 by TBATS model. 
```{r}
covid_ts_forecast(start_date='2020-04-30', fvar="new_case", model=c("tbats"), pred.days = 30)
```

* Forecasting new COVID-19 deaths in the states MA, CA, NY, NJ, IL and TN from now until next 7 days using data from June 01, 2020 by ARIMA model. 
```{r}
covid_ts_forecast(start_date='2020-06-01', fvar="new_death", state.filter=c("MA","CA","NY","NJ","IL","TN"), model=c("auto.arima"), pred.days = 7)
```

* Forecasting new COVID-19 deaths across Midwest USA and West USA from now until next 20 days using data from Jan 22, 2020 by ARIMA model. Here model default is ```auto.arima``` and start date default is ```2020-01-22```
```{r}
covid_ts_forecast(fvar="new_death",region.filter=c("Midwest","West"),pred.days=20)
```


